---
title: "Estimating concentration indices with conindex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating concentration indices with conindex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(conindex)
```

## Overview

`conindex` computes rank-dependent inequality indices — the family of
concentration indices used in health economics to measure socioeconomic
inequality in an outcome. It is a faithful R translation of the Stata command of
the same name (O'Donnell et al. 2016), reproducing its index values and standard
errors.

There is **no default index**: options select which index is computed.

## A worked example on synthetic data

The real examples in O'Donnell et al. (2016) use Demographic and Health Survey
data that cannot be redistributed. Here we use a small synthetic dataset with an
outcome `h` and a socioeconomic ranking variable `ses`.

```{r}
set.seed(1)
n <- 800
df <- data.frame(
  ses = rnorm(n),
  wt  = runif(n, 0.5, 1.5)
)
df$h    <- pmax(0.05, 2 + 0.6 * df$ses + rnorm(n))   # ratio-scale outcome
df$hbin <- as.numeric(df$h > median(df$h))            # bounded [0, 1] outcome
```

### Standard concentration index

```{r}
conindex(df, "h", rankvar = "ses", weights = "wt", truezero = TRUE)
```

### Gini coefficient

Ranking by the outcome itself yields a Gini coefficient:

```{r}
conindex(df, "h", weights = "wt", truezero = TRUE)
```

### Generalized concentration index

```{r}
conindex(df, "h", rankvar = "ses", weights = "wt", truezero = TRUE,
         generalized = TRUE)
```

### Bounded variables: Wagstaff and Erreygers

For a bounded (e.g. binary) outcome, specify `bounded` and `limits`:

```{r}
conindex(df, "hbin", rankvar = "ses", bounded = TRUE, limits = c(0, 1),
         erreygers = TRUE)
conindex(df, "hbin", rankvar = "ses", bounded = TRUE, limits = c(0, 1),
         wagstaff = TRUE)
```

### Distributionally sensitive indices

The extended index (`v`) and symmetric index (`beta`) allow alternative attitudes
to inequality. With `v = 2` (or `beta = 2`) they coincide with the standard CI.
Standard errors are not produced for these indices.

```{r}
conindex(df, "h", rankvar = "ses", truezero = TRUE, v = 1.5)
conindex(df, "h", rankvar = "ses", truezero = TRUE, beta = 1.5)
```

### Robust and clustered standard errors

```{r}
df$psu <- sample(1:40, n, replace = TRUE)
conindex(df, "h", rankvar = "ses", truezero = TRUE, cluster = "psu")
```

### Comparing groups

`compare` computes the index within each group and tests homogeneity with a
small-sample F-test and (for two groups) a large-sample z-test.

```{r}
df$urban <- rbinom(n, 1, 0.4)
conindex(df, "h", rankvar = "ses", truezero = TRUE, compare = "urban")
```

## Interactive use

`run_conindex_app()` opens a Shiny GUI for point-and-click estimation.

## References

O'Donnell, O., O'Neill, S., Van Ourti, T. & Walsh, B. (2016). conindex:
Estimation of concentration indices. *The Stata Journal*.
